---
title: "Statistical Thinking in Biology Research"
subtitle: "Probability and Statistical Inference"
author: Terry Neeman
institute: Australian National University
date: "30th July 2020"
fontsize: 12pt
mathspec: true # for metropolis theme
aspectratio: 169 # 16:9
#classoption: handout
output:
 beamer_presentation:
   theme: "Dresden" 
   # https://deic-web.uab.cat/~iblanes/beamer_gallery/index_by_theme.html
   #theme: "metropolis" # special
   colortheme: "dolphin"
   # https://deic-web.uab.cat/~iblanes/beamer_gallery/index_by_color.html
   fonttheme: "structurebold"
   # https://deic-web.uab.cat/~iblanes/beamer_gallery/index_by_font.html
   latex_engine: xelatex # for metropolis theme
   toc: false
   slide_level: 2
   keep_tex: true
   #includes:
   #   in_header: preamble.tex
header-includes:
  - \usepackage{fancyvrb}
---
 
## A few key ideas

* Probability: understanding possible outcomes under a set of "rules"
* Domain of probability: mathematics ("theoretical", "proof")
* Statistics: Given a set of outcomes, what can we *infer* about the possible rules?
* Domain of statistics: real world data ("pragmatic", "heuristic")
 
::: block
### Probability and Statistics are two sides of the same subject.   
:::

 
## Probability: measures of uncertainty

* Sample space: space of possible outcomes 
* Distribution: relative frequencies (probabilities) of each outcome
* Summaries of distributions: average (expected) outcome, variation around average
 
## Examples of common distributions in biological research

* Normal distribution
   * family of distributions
   * sample space $(-\infty, \infty)$
   * defined by two parameters: mean and standard deviation (variance)
   * many biological measures normally distributed, e.g. height, weight

      

## Sample from a normal distribution

```{r, message=FALSE, fig.height=1.5, fig.width=2,fig.align = 'center'}
library(tidyverse)
sample_normal <- tibble(x = rnorm(n=1e5, mean = 5, sd = 2))
ggplot(sample_normal, aes(x = x))+
   geom_density()
```


## Plot the THEORETICAL Normal distribution
```{r, message=FALSE, fig.height=1.5, fig.width=2,fig.align = 'center'}
outcomes<-seq(-5,15, length.out = 1e4)
out_normal <- tibble(outcomes = outcomes, 
                     rel_freq = dnorm(outcomes, mean = 5, sd = 2))
ggplot(out_normal, aes(x=outcomes,y = rel_freq))+
   geom_line()
```

## Examples of common distributions in biological research

* Binomial distribution
   * family of distributions
   * Describes potential outcomes: #successes out of n independent trials
   * defined by two parameters: 
      * n = # of independent trials
      * p = probability of success in a trial
      
   * sample space ${0, 1, ..., n}$

## Sample from a binomial distribution

```{r, message=FALSE, fig.height=1.5, fig.width=2, fig.align = 'center'}
sample_binomial <- tibble(x = rbinom(1e4, size = 100, prob = 0.95))
ggplot(sample_binomial, aes(x = x))+
   geom_histogram(binwidth = 0.5)
```


## Plot the THEORETICAL binomial  distribution
```{r, message=FALSE, fig.height=1.5, fig.width=2, fig.align = 'center'}
outcomes<-seq(71,100, by=1)
outcomes_binomial <- tibble(outcomes = outcomes, 
                     prob = dbinom(outcomes, size=100, prob=0.95))
ggplot(outcomes_binomial, aes(x=outcomes,y = prob))+
   geom_bar(stat="identity")
```

## Sampling from a distribution: A data-generating machine

```{r normal_machine,fig.height=3, fig.width=4, fig.align = 'center', echo=FALSE, warning=FALSE}
library(ggforce)
dat = data.frame(x0=c(.25,.25,.60), y0=c(.25,.55,.25), r=.1)
ggplot() + scale_x_continuous(limits = c(0,1)) +
  scale_y_continuous(limits = c(0,1))+
  geom_circle(aes(x0=x0, y0=y0,r=r),data=dat)+
  annotate("rect", xmin = .15, xmax = .35, ymin = .15, ymax = .35,
           alpha = .2, fill="blue", colour= "black")+
  geom_segment(aes(x=.25, y=.45, xend=.25, yend=.35), 
               arrow=arrow(), size=1, color="blue")+
  geom_segment(aes(x=.5, y=.25, xend=.35, yend=.25), 
               arrow=arrow(), size=1, color="blue")+
  annotate("text", x=.25, y=.25, label="Y[{i}]", parse = TRUE,size=5)+
  annotate("text", x=.6, y=.25, label="\u03c3 (SD)", size=3)+
  annotate("text", x=.25, y=.55, label="\u03BC (mean)", size=3)+
  #ggtitle("Graphical model for normal distribution") +
  theme_void()

```

Now let's look at this from the other end. Using data, can we build a machine
that may have generated our data?

## Machine 1: Precipitation -> Yield
```{r, fig.height=1.4, fig.width=2, fig.align = 'center'}
set.seed(202073)
annual_rain<-seq(11,100, 1)
yield <- 2 + 5*annual_rain - 0.04* annual_rain^2 + rnorm(90,0,10)
yield_dat<-tibble(annual_rain = annual_rain, yield=yield)
ggplot(yield_dat, aes(annual_rain, yield))+geom_point()
```

## Draw a picture of Machine 1


## Machine 2: Precipitation -> Yield, AR process
```{r}
set.seed(202073)
annual_rain<- rep(30,50); yield <-rep(150,50)
for (i in 2:50) {
     annual_rain[i]<- 0.3*annual_rain[i-1] + rnorm(1,30,10)
     yield[i]<- 5*annual_rain[i] - 0.04* annual_rain[i]^2 + 
        0.2*yield[i-1] + rnorm(1,0,10)
}
yield_dat2<-tibble(year = 1951:2000, rain = annual_rain, yield=yield)
```

## Draw a picture of Machine 2


## Machine 2: Precipitation -> Yield, AR process
### Annual rain between 1951 and 2000
```{r, fig.height=1.8, fig.width=3.3, fig.align = 'center'}
ggplot(yield_dat2, aes(year, rain))+geom_line()
```

## Machine 2: Precipitation -> Yield, AR process
### Crop yield between 1951 and 2000
```{r, fig.height=1.8, fig.width=3.3, fig.align = 'center'}
ggplot(yield_dat2, aes(year, yield))+geom_line()
```

## Machine 2: Precipitation -> Yield, AR process
### Crop yield vs Annual rain
```{r, fig.height=1.8, fig.width=3.3, fig.align = 'center'}
ggplot(yield_dat2, aes(rain, yield))+geom_point()
```

## Summary

* A probability distribution: a set of possible outcomes and associated probabilities
* Data generating process: set of rules for generating set of outcomes
* Probability: from rules to data
* Statistics:  from data to rules

::: block
### Statistics: re-constructing the rules, given the data 
### The Ultimate Challange!
:::